# Ångest

## Läkemedelsbehandling indikationer

Se [@Garakani2020]

```{r anxiety-medications } 

tbl <- fread('../notes/anxiety_medications.csv')
## i stort sett alla som är godkända i EU är det också i USA

## antidepressiva
tbl1 <- tbl[type %in% c("ssri", "snri", "other_antidepressant", "tca", "maoi"), ]
tbl1[tbl1 == "EU, USA"] <- "EU"
tbl1[tbl1 == "USA"] <- "EU"
tbl1[tbl1 == "EU"] <- "Ja"
tbl1[, type := NULL]

kable(tbl1, caption = "Antidepressiv medicinering med godkännande i EU eller USA mot ångestsyndrom")

## antipsykotiska medel

tbl1 <- tbl[type %in% c("antipsychotics"), ]

## i stort sett alla som är godkända i EU är det också i USA
tbl1[tbl1 == "EU, USA"] <- "EU"
tbl1[tbl1 == "USA"] <- "EU"
tbl1[tbl1 == "EU"] <- "Ja"
tbl1[, type := NULL]

kable(tbl1, caption = "Antipsykotisk medicinering med godkännande i EU eller USA mot ångestsyndrom")

tbl1 <- tbl[type %in% c("other"), ]
tbl1[tbl1 == "EU, USA"] <- "EU"
tbl1[tbl1 == "USA"] <- "EU"
tbl1[tbl1 == "EU"] <- "Ja"
tbl1[, type := NULL]

kable(tbl1, caption = "Annan medicinering med godkännande i EU eller USA mot ångestsyndrom")

tbl1 <- tbl[type %in% c("benzodiazepines"), ]
tbl1[tbl1 == "EU, USA"] <- "EU"
tbl1[tbl1 == "USA"] <- "EU"
tbl1[tbl1 == "EU"] <- "Ja"
tbl1[, type := NULL]

kable(tbl1, caption = "Bensodiazepiner med godkännande i EU eller USA mot ångestsyndrom")

``` 

Abbreviations: GAD: generalised anxiety disorder; MAOIs: monoamine oxidase inhibitors; OCD: obsessive compulsive disorder; PD: panic disorder; PTSD: post-traumatic stress disorder; SAD: social anxiety disorder. 

>The off-label use of the listed antidepressants, antipsychotics and others that are not approved for any anxiety, trauma-related or obsessive disorder is supported by clinical studies, meta-analysis and more recent network meta-analysis [@Garakani2020].


### Ångestsyndrom antidepressiva

Med reservation för publikationsbias (positiva resultat korrekt ca 5 gånger mer än negativa[@Roest2015]) tycks SSRI enligt en metaanalys (K = 57, N = 16056) ha effekt mot ångestsyndrom (SMD = 0.60; 95% CI [0.55–0.65], z = 29.2, P < 0.001, vs placebo) [@Jakubovski2019]. Det mesta av effekten kommer inom 1-2 månader med viss ytterligare effektökning upp till kanske tre månader. Effektstorleken baserat på FDAs register är ca 0.33 (Hedges g, liten effekt).

Över tid föreligger små eller inga skillnader i respons beroende på diagnos (SA, GAD, PD) för SSRI. För SNRI kan PD möjligen svara något mindre på behandlingen än de båda andra diagnoserna[@Jakubovski2019]. 

En minimalt dos-respons-samband (SMD 0.5 vs 0.6) kan finnas i intervallet 50-300 mg imipraminekvivalenter (7.5-23 mg escitalopram) för SSRI men ej för SNRI. Högre dos ökar också biverkningarna[@Jakubovski2019].

SSRI kan möjligen vara något mer effektivt än SNRI (SMD 0.6 vs 0.5), men ej vid "longitudinell analys". Skillnader mellan olika preparat inom SSRI respektive SNRI-gruppen kunde ej påvisas[@Jakubovski2019].

> 100 mg of imipramine = 120 mg of sertraline = 100 mg of fluvoxamine = 20 mg of paroxetine or fluoxetine = 30 mg of citalopram = 15 mg of escitalopram = 100 mg of venlafaxine = 40 mg of duloxetine.


<!-- [@Beyer2024] -->


<!-- Comprehensive, evidence-based guides to direct treatment of generalized anxiety disorder are limited, with one noteworthy exception being the “Canadian Clinical Practice Guidelines for the Management of Anxiety, Posttraumatic Stress and Obsessive-Compulsive Disorders” (45). In this sizeable review, the authors stratified treatment interventions according to established efficacy, quality of evidence, and tolerability and safety considerations; furthermore, they proposed a hierarchy of first-, second-, and third-line monotherapy options as well as ranked adjunctive treatment options. The authors concluded that SSRIs and SNRIs along with pregabalin should be considered first-line treatments. They characterized TCAs, BZDs, vortioxetine, bupropion XL, buspirone, quetiapine, and hydroxyzine as second-line treatments. They qualified miscellaneous antidepressants and valproic acid (VPA) in rare formulations as third-line treatments. Finally, for adjunctive treatment, they favored pregabalin over second-generation antipsychotics (45). https://focus.psychiatryonline.org/doi/10.1176/appi.focus.20200048 -->

<!-- Se /notes/antianxiety-medication-Katzman2014.csv -->
<!-- Se[@Katzman2014] -->
<!-- MAOI = monoamine oxidase inhibitor; RIMA = reversible inhibitor of monoamine oxidase A; SNRI = serotonin–norepinephrine reuptake inhibitor; SR = sustained release; SSRI = selective serotonin reuptake inhibitor; TCA = tricyclic antidepressant; XR = extended release  -->
<!-- *Monotherapy and augmentation -->
<!-- †Augmentation only, usually added to an antidepressant medication -->
<!-- §Two studies: combination of exposure therapy and clomipramine; another an augmentation study adding clomipramine to fluoxetine -->
<!-- ‡Used as monotherapy and in one study as an augmentation agent to mirtazapine -->
<!-- First-line	Level 1 or Level 2 evidence plus clinical support for efficacy and safety -->
<!-- Second-line	Level 3 evidence or higher plus clinical support for efficacy and safety -->
<!-- Third-line	Level 4 evidence or higher plus clinical support for efficacy and safety -->
<!-- Not recommended	Level 1 or Level 2 evidence for lack of efficacy -->


### Ångestsyndrom antidepressiva 1:a/2:a-handsval

```{r } 
#| label: anxiety-can 

dta <- fread("../notes/antianxiety-medication-Katzman2014.csv", na.strings = c(""))
dta[, drug:=str_squish(drug)]
dta[, drug:=str_trim(drug)]
dta[, drug := gsub("[^A-z| ]", "", drug)]

## labels add
lookup <- dta[!is.na(class), .(class, subclass, drug)]
setkey(dta, class)
setkey(lookup, class)
dta <- lookup[dta, on = c("drug")]
dta[is.na(class), class := i.class]
dta[is.na(subclass), subclass := i.subclass]
## dta[is.na(class), ] ## check
dta[, i.subclass:=NULL]
dta[, i.class:=NULL]

## tmp[is.na(mono), mono := ""]
## tmp[is.na(adj), adj := ""]

## tmp <- dta[mono==1|mono==2|adj==1|adj==2, .(drug, indication, mono, adj)]
## tmp[, mono_adj := paste0("m", mono, "a", adj)]
## tmp[, mono_adj := gsub("mNA", "", mono_adj)]
## tmp[, mono_adj := gsub("aNA", "", mono_adj)]

## Monoterapier
tmp <- dta[mono==1|mono==2, .(drug, indication, mono, class)]
tmp <- unique(tmp)
tbl <- dcast(drug+class~indication, data = tmp, value.var="mono")

## Curate paroxetine IR/CR
tbl[drug == "Paroxetine", sa := 1]
tbl <- tbl[drug!="Paroxetine CR" & drug!="Paroxetine IR", ]

setorder(tbl, class, gad, ocd, pd, ptsd, sa, na.last=TRUE)
tbl[, drug := gsub(" XR| XL| CR| IR", "", drug)]
tbl <- unique(tbl)
toremove <- c("Fluvoxamine", "Phenelzine", "Imipramine", "Bromazepam")
tbl <- tbl[!drug %in% toremove, ]

colnames(tbl) <- toupper(colnames(tbl))
mycols <- colnames(tbl)
mycols <- mycols[mycols != "CLASS"]
tbl_ad <- tbl[CLASS == "antidepressant", ..mycols]
tbl_other <- tbl[CLASS != "antidepressant", ..mycols]

## Adjunctive treatments
tmp <- dta[adj==1|adj==2, .(drug, indication, adj, class)]
tbl <- dcast(drug+class~indication, data = tmp, value.var="adj")
colnames(tbl) <- toupper(colnames(tbl))
mycols <- colnames(tbl)
mycols <- mycols[mycols != "CLASS"]
tbl_adj <- tbl[, ..mycols]

## Kable
kable(tbl_ad)

``` 

1:a och 2:a hands val monoterapi.

Ref [@Katzman2014]

### Ångestsyndrom andra läkemedel

```{r } 
#| label: anxiety-meds-other
kable(tbl_other)

```

1:a och 2:a hands val monoterapi [@Katzman2014].

### Ångestsyndrom tilläggspreparat

```{r } 
#| label: anxiety-meds-adjunctive
setorder(tbl_adj, GAD, OCD, PD, PTSD, na.last=TRUE)
kable(tbl_adj)
``` 
1:a och 2:a hands val som tillägg till andra läkemedel.

Ref [@Katzman2014]



## Ångestsyndrom hos barn

Tre relaterade ångestsyndromen hos barn är generaliserat ångestsyndrom, social ångest och separationsångest. Förekomsten av dessa medför ökad risk för liknande problem, samt tillkomst av depression och drog/alkoholberoende [@Wehry2015].

Spontan remission förekommer, men oftast är besvären persisterande med varierande svårighetsgrad över tid. De ändrar ibland karaktär under utvecklingsåren och senare under livet. Faktorer kopplade till sämre prognos är kvinnligt kön, högre ålder, förstagradssläkting med ångestsyndrom, högre symtomnivå vid diagnos, och flera samtidiga ångestsyndrom. Bra familjefungerande kan vara skyddande [@Wehry2015].

Både psykoterapi (tex KBT) och läkemedel (SSRI inklusive sertralin, paroxetin och fluoxetin; SNRI inklusive venlafaxin, TCA) har effekt. Kombinationsbehandling med läkemedel och psykoterapi har i vissa studier varit mer effektivt än endera. Atomoxetin har effekt mot ångest vid samtidig ADHD. Buspiron har ej visat effekt mot ångest hos barn, ej heller bensodiazepiner. Optimal dosering och behandlingslängd är okänd [@Wehry2015].

Spence Children’s Anxiety Scale kan användas för screening, stöd för diagnostik, och att följa behandlingseffekt. En svensk version finns på sfbup.

Se Assessment and Treatment of Anxiety Disorders in Children and Adolescents Anna M. Wehry [@Wehry2015] 

<!-- ## Epidemiology -->

<!-- 20% av barn och unga har ångestsyndrom. Associerat med ökad risk för senare problem inkl ångestsyndrom, drogberoende, och depression. -->

<!--  Ofta samsjuklighet mellan social ångest, generaliserad ångest, och separationsångest. Kan ses som en "triad" med liknande utveckling, respons på farmakologisk och psykoterpaeutisk behandling. -->

<!-- Spontan remission förekommer, men vanligen persisterande med uppgång och nedgång och att bevsären ändrar karaktär under utvecklingen och senare under livet. -->

<!-- Faktorer associerade m sämre prognos: Högre ålder, förstagradssläkting m ångestsyndrom, högre symtomnivå vid diagnos, flera komorbiditeter. -->

<!-- Faktorer associerade m bättre prognos: Manligt kön, bättre familjefungerande, -->

<!-- ## Neurobiology -->

<!-- Flera nätverk och strukturer implicerade. Amygdala, cingulate cortex, och VLPFC kan ha ventrala roller.  -->

<!-- ## Rating scales -->

<!-- Several well-studied child self-report screening measures for anxiety exist and may be used in patients >8 years of age. First, the Multidimensional Anxiety Scale for Children [33], and the Screen for Child Anxiety and Related Emotional Disorders (SCARED) [34], and the Spence Children’s Anxiety Scale (SCAS) [35] were developed to be sensitive and specific for assessing anxiety in youth and are useful in clinical practice to monitor treatment progress [36]. The Preschool Anxiety Scale is a parent report adapted from the SCAS that was developed for screening for anxiety in young children (ages 2.5 to 6.5). The SCARED and SCAS are available online/free access. Second, the Pediatric Anxiety Rating Scale, which is frequently utilized as a primary or secondary outcome measure in psychopharmacologic trials of youth, is a clinician-rated measure used to assess the severity of anxiety symptoms and change over time [37]. Third, with regard to social phobia or social anxiety, the Social Anxiety Scale, the Social Worries Questionnaire, and the social phobia subscale of SCARED are brief screening measures for social phobia/social anxiety symptoms. Fourth, a set of novel dimensional anxiety scales was recently generated during the DSM revision process to support clinical decision-making and monitoring treatment progress (www.psychiatry.org/dsm5). These self-rated scales can be used for children aged 11 or older and comprise the core constructs of fear and anxiety together with cognitive, physical and behavioral symptoms in a consistent and brief way, using a common template for each of the anxiety disorders -->

<!-- Dimensional scales, se: https://onlinelibrary.wiley.com/doi/full/10.1002/mpr.1369 -->

<!-- Skattningsskalor SFBUP (SCAS): http://www.sfbup.se/vard/skattningsskalor/ -->

<!-- ## Treatment -->

<!-- In a large, multisite study of youth with moderate to severe GAD, SoP and SAD, the Child/Adolescent Anxiety Multimodal Study (CAMS), 488 children and adolescents (aged 7-17 years) were randomized to one of three treatment groups (sertraline monotherapy, cognitive behavioral therapy [CBT], or sertraline + CBT) for 12 weeks [7]. In terms of clinical improvement and symptom severity, all treatment groups were superior to placebo (24%), and the combination therapy (sertraline + CBT) was significantly more efficacious (81%) than either group treated with either medication (55%) or CBT (60%) alone. the naturalistic 6-year follow-up study of this sample (n=288) revealed that these treatment effects were durable and sustained, with remission rates of 48.8%, 51.9%, and 45.8% for combination therapy, sertraline only, and CBT only, respectively.  Alternatively, nearly half of the acute responders relapsed during the follow-up period. -->

<!-- Social Effectiveness Therapy for Children (SET-C) versus fluoxetine or placebo in children and adolescents aged 7-17 with social phobia (n=122). SET-C was more effective than placebo on primary outcome measures and specific symptom measures. Additionally, SET-C was superior to fluoxetine in terms of rate of treatment response (79% versus 36.4%, p<0.001), lack of posttreatment diagnosis, and higher end-state functioning [39].  -->

<!-- SSRI (inkl Sertralin, fluoxetin) och SNRI (venlafaxin, paroxetin mm), och TCA har effekt. -->

<!-- Atomoxetine has been evaluated in youth with ADHD and a co-occurring anxiety disorders (SAD, GAD and/or SoP; PARS score ≥15) in a 12-week, multicenter, double-blind, placebo-controlled study [70]. Patients (aged 8-17 years, mean age: 12.0±3.0 years, N=176) were treated with atomoxetine which was titrated to 1.8 mg/kg/day (mean dose: 1.3 mg/kg/day). Reductions in both anxiety and ADHD symptoms were observed with atomoxetine treatment and this SNRI was well-tolerated, although decreased appetite was more frequently observed in atomoxetine-treated patients -->

<!-- Relatedly, few studies have examined optimal “dosing” of either psychotherapeutic or psychopharmacologic treatments and little is known regarding next-step interventions for youth with anxiety disorders, who do not response to initial evidence-based strategies. Finally, the optimal duration of treatment for pediatric patients with anxiety disorders is generally unknown, with both psychopharmacolgic and psychotherapeutic studies being frequently focused on short-term, acute effects. -->
